Estimating duration in the presence of misclassification

### 2024-03-31 cis-imperfect-testing.tex
proof ch 6

In this section, I introduce a simple model of false negatives (\ie allowing $\psens < 1$) into the posterior density derived in \cref{E-perf-test:sec:model}.
Using a simple model, in particular a constant test sensitivity, means calculating likelihood remains tractable.

@@

To formalise this, I define analogous variables to those in \cref{E-perf-test:sec:posterior}.
To start with, define the relevant testing times for episodes with observations $\vec{\nu}_k$ as $\sched'_k = \{ t \in \sched_{i_k} \ssep r_k^{(b)} \leq t \leq r_k^{(e)} + 1 \}$ and let $m_k$ denote the size of this set.
Denote the elements of $\sched'_k$ by $t_{k,1} < \dots < t_{k,m_k}$.
Let $\set{E}' = \{ \vec{\nu}_1', \dots, \vec{\nu}_{N_E'}' \}$ be a new set of possible observations of detected episodes, including the pattern of positive and negative tests.
Represent the observations such that for any $\vec{\nu}_k' \in \set{E}'$, $\vec{\nu}_k' = [\vec{\nu}_{k}, \vec{y}_k]^T$, where $\vec{y}_j \in \{ 0, 1 \}^{m_j}$ with $y_{j,l}$ being the test result at time $t_{j,l}$.

@@

I will modify $p_{ik}$ to form $p_{ik}' = \prob(O'(j) = \vec{\nu}'_k \mid i(j) = i_k, \vec{\theta})$, taking into account false negatives.
I consider a mixture of two scenarios, whether the final test in $\sched'_{i(j)}$ is a false negative, with the mixture probability determined by the test sensitivity.

@@

First, the case when $E_j \leq r_k^{(e)}$.
In this case, the test at $r_k^{(e)}+1$ is a true negative and the end of the episode is interval censored as in the previous chapter.
% In this case, the test at $r_k^{(e)} + 1$ is a true negative, as are all other tests not in $\sched'_{i(j)}$.
The true negative occurs with probability 1, by the assumption of no false positives.
\begin{align}
&\prob(\vec{\nu}_k, E_j \leq r_k^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\
&= \prob(\vec{\nu}_k, l_k^{(e)} \leq E_j \leq r_k^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\ % &\text{as no false positives}
&= \prob(\vec{\nu}_k \mid l_k^{(e)} \leq E_j \leq r_k^{(e)}, B_j = b, p_\text{sens}, \vec{\theta}) \prob(l_k^{(e)} \leq E_j \leq r_k^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\
&= p_\text{sens}^{t_+} (1 - p_\text{sens})^{f_-} \left( S_{\vec{\theta}}(l_k^{(e)} - b + 1) - S_{\vec{\theta}}(r_k^{(e)} - b + 2) \right)
\label{imperf-test:eq:ll-ei-lt-ri}
\end{align}

Second, the case when $E_j > r_k^{(e)}$.
In this case, the test at $r_k^{(e)}$ is a false negative, occurring with probability $(1 - p_\text{sens})$.
To avoid having to consider tests after $r_k^{(e)}$, which could greatly complicate the likelihood, I model this case as the episode being right censored at $r_k^{(e)}$.
Taking the same approach as before:
\begin{align}
&\prob(\vec{\nu}_k, E_j > r_k^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\
&= \prob(\vec{\nu}_k \mid E_j > r_k^{(e)}, B_j = b, p_\text{sens}, \vec{\theta}) \prob(E_j > r_k^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\
&= p_\text{sens}^{t_+} (1 - p_\text{sens})^{f_-} (1 - p_\text{sens}) S_{\vec{\theta}}(r_k^{(e)} - b_j + 2)
\label{imperf-test:eq:ll-ei-gt-ri}
\end{align}

These expressions can now be used to form the full replacement for $p_{ik}$, $p'_{ik} = \prob(O'(j ) = \vec{\nu}_k' \mid i(j) = i_k)$.
First, augment the data with $b_j$, and split into the cases just discussed:
\begin{align}
p_{ik}'
=& \prob(\vec{\nu}_k \mid p_\text{sens}, \vec{\theta}) \\
=& \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \left( \prob(\vec{\nu}_k, E_j \leq r_k^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) + \prob(\vec{\nu}_k, E_j > r_k^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \right) \\
  &\times \prob(B_j = b \mid p_\text{sens}, \vec{\theta}).
\intertext{Now, substitute in \cref{imperf-test:eq:ll-ei-lt-ri,imperf-test:eq:ll-ei-gt-ri} and take out the common factor:}
=\ &  p_\text{sens}^{t_+} (1 - p_\text{sens})^{f_-} \\
 & \times \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \left( S_{\vec{\theta}}(l_k^{(e)} - b + 1) - S_{\vec{\theta}}(r_k^{(e)} - b + 2) + (1 - p_\text{sens}) S_{\vec{\theta}}(r_k^{(e)} - b + 2) \right) \\ 
  & \times \prob(B_j = b \mid p_\text{sens}, \vec{\theta}) \\
=\ &  p_\text{sens}^{t_+} (1 - p_\text{sens})^{f_-} \\
  & \times \sum_{b_j = l_k^{(b)}}^{r_k^{(b)}} \left( S_{\vec{\theta}}(l_k^{(e)} - b_j + 1) - p_\text{sens} S_{\vec{\theta}}(r_k^{(e)} - b_j + 2) \right) \\
  & \times \prob(B_j = b \mid p_\text{sens}, \vec{\theta}).
\label{imperf-test:eq:pia-prime}
\end{align}
Note that if $p_\text{sens} = 1$ then $p_{ik}' = p_{ik}$ (see \cref{perf-test:eq:pia}).

If $\psens$ is fixed (\ie has a point prior) and $\prob(B_j = b \mid \psens, \vec{\theta}) \propto 1$ (as assumed in \cref{E-perf-test}) then:
\begin{align}
p_{ik}'
&\propto \sum_{b = l_k^{(b)}}^{r_k^{(b)}} S_{\vec{\theta}}(l_k^{(e)} - b_j + 1) - p_\text{sens} S_{\vec{\theta}}(r_k^{(e)} - b + 2).
\label{imperf-test:eq:pia-prime-constant}
\end{align}

@@

This false negative would occur at the first test after the infection episode begins, on day $b_j + \tau_{\sched_{i(j)}}(b_j)$ (see \cref{E-perf-test:sec:prob-undetected}).
A false negative occurring requires that the episode has not yet ended but a negative still occurs.
The episode has not yet ended at the time of the test if $e_j = b_j + d_j - 1 \geq b_j + \tau_{\sched_{i(j)}}(b_j) + 1$, that is the duration of the infection $d_j \geq \tau_{\sched_{i(j)}}(b_j) + 1$.
Conditional on the episode having not yet ended, the test result is negative with probability $1 - \psens$.

For there to be no subsequent positive tests, all tests up until day $e_j$ are false negatives.
I assume there is a negligible probability of two false negatives.
Therefore, this can only occur if the episode ends before another test.
Denote the number of days between $b_j$ and the test following the false negative as $\tau^2_{\sched_{i(j)}}(b_j) \stackrel{\text{def}}{=} \tau_{\sched_{i(j)}}(\tau_{\sched_{i(j)}}(b_j) + 1)$.
Then, the episode ends before this test if $d_j \leq \tau^2_{\sched_{i(j)}}(b_j) + 1$.

Therefore, this mechanism causes episode $j$ to be undetected if all the following conditions hold.
\begin{enumerate}
    \item The episode would have been detected considering only the mechanisms in \cref{E-perf-test:sec:prob-undetected}. That is $\min(\sched_{i(j)}) < b_j \leq T_{i(j)}$ and $e_j \geq \tau_{\sched_{i(j)}}(b_j) + b_j$.
    \item The episode ends in the interval $[\tau_{\sched_{i(j)}}(b_j) + b_j, \tau^2_{\sched_{i(j)}}(b_j) + b_j]$.
      Note that the lower bound here is exactly the bound on $e_j$ in the previous condition.
      Equivalently, $\tau_{\sched_{i(j)}}(b_j) + 1 \leq d_j \leq \tau^2_{\sched_{i(j)}}(b_j) + 1$.
    \item A false negative occurs on day $\tau_{\sched_{i(j)}}(b_j) + b_j$. Conditional on the previous condition, this occurs with probability $1 - \psens$.
\end{enumerate}

The probability of this occurring, conditional on $B_j = b$ where $\min(\sched_{i(j)}) < b \leq T_{i(j)}$ is:
\begin{align}
&\prob \left(
    \tau_{\sched_{i(j)}}(b) + 1 \leq D_j \leq \tau^2_{\sched_{i(j)}}(b) + 1
    \mid B_j = b, \vec{\theta} \right) (1 - \psens) \\
&= \left( S_{\vec{\theta}}(\tau_{\sched_{i(j)}}(b) + 1) - S_{\vec{\theta}}(\tau^2_{\sched_{i(j)}}(b) + 2) \right) (1 - \psens).
\end{align}
Summing over $b$, in the same way as \cref{perf-test:eq:piu}, gives:
\begin{align}
\zeta = (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(\sched_{i(j)}) + 1}^{T_{i(j)}} \left( S_{\vec{\theta}}(\tau_{\sched_{i(j)}}(b) + 1) - S_{\vec{\theta}}(\tau^2_{\sched_{i(j)}}(b) + 2) \right).
\label{imperf-test:eq:zeta}
\end{align}

$p_{iu}'$ is the probability of episode $i$ being undetected, considering both the mechanisms from \cref{E-perf-test} and the new mechanism here.
These mechanisms are mutually exclusive.
Hence, $p_{iu}'$ is the sum of these, $p_{iu}' = p_{iu} + \zeta$.
As in \cref{E-perf-test:sec:prob-undetected}, $1 - p_{iu}'$ is the required quantity.
\begin{align}
1 - p_{iu}'
=& 1 - p_{iu} - \zeta \\
=& \frac{1}{T} \sum_{b = \min \sched_{i} + 1}^{T_{i(j)}} S_{\vec{\theta}}(\tau_{\sched_{i}}(b) + 1) \\
&- (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(\sched_{i(j)}) + 1}^{T_{i(j)}} \left( S_{\vec{\theta}}(\tau_{\sched_{i(j)}}(b) + 1) - S_{\vec{\theta}}(\tau^2_{\sched_{i(j)}}(b) + 2) \right) \\
=& \frac{1}{T} \sum_{b=\min(\sched_{i(j)}) + 1}^{T_{i(j)}} \left( p_\text{sens} S_{\vec{\theta}}(\tau_{\sched_{i(j)}}(b) + 1) + (1 - p_\text{sens}) S_{\vec{\theta}}(\tau^2_{\sched_{i(j)}}(b) + 1)\right).
\label{imperf-test:eq:pit-prime}
\end{align}

>>>>

In this section, I introduce a simple model of false negatives (\ie allowing $\psens < 1$) into the posterior density derived in \cref{E-perf-test:sec:model}.
Using a simple model, in particular a constant test sensitivity, means calculating the likelihood remains tractable.

@@

To formalise this, I define analogous variables to those in \cref{E-perf-test:sec:posterior}.
To start with, define the relevant testing times for any episode $j$ with $O(j) = \vec{\nu}_k$ as $\sched'_k = \{ t \in \sched_{i_k} \ssep r_k^{(b)} \leq t \leq r_k^{(e)} + 1 \}$ and let $m_k$ denote the size of this set.
Denote the elements of $\sched'_k$ by $t_{k,1} < \dots < t_{k,m_k}$.
Let $\set{E}' = \{ \vec{\nu}_1', \dots, \vec{\nu}_{N_E'}' \}$ be a new set of possible observations of detected episodes, including the pattern of positive and negative tests.
Represent the observations such that for any $\vec{\nu}_k' \in \set{E}'$, $\vec{\nu}_k' = [\vec{\nu}_{k}, \vec{y}_k]^T$, where $\vec{y}_k \in \{ 0, 1 \}^{m_k}$ with $y_{k,l}$ being the test result at time $t_{k,l}$.

@@

I will modify $p_{ik}$ to form $p_{ik}' = \prob(O'(j) = \vec{\nu}'_k \mid i(j) = i_k, \vec{\theta})$, taking into account false negatives.
I consider a mixture of two scenarios, whether the final test in $\sched'_{k}$ is a false negative, with the mixture probability determined by the test sensitivity.

@@

First, the case when $E_j \leq r_k^{(e)}$.
In this case, the test at $r_k^{(e)}+1$ is a true negative and the end of the episode is interval censored as in the previous chapter.
% In this case, the test at $r_k^{(e)} + 1$ is a true negative, as are all other tests not in $\sched'_{k}$.
The true negative occurs with probability 1, by the assumption of no false positives.
\begin{align}
&\prob(O'(j) = \vec{\nu}_k', E_j \leq r_k^{(e)} \mid B_j = b, i(j) = i_k, \psens, \vec{\theta}) \\
&= \prob(O'(j) = \vec{\nu}_k', l_k^{(e)} \leq E_j \leq r_k^{(e)} \mid B_j = b, i(j) = i_k, \psens, \vec{\theta}) &\text{the test at $l_k^{(e)}$ is positive} \\
&= \prob(O'(j) = \vec{\nu}_k' \mid l_k^{(e)} \leq E_j \leq r_k^{(e)}, B_j = b, i(j) = i_k, \psens, \vec{\theta}) \\
&\ \ \  \times \prob(l_k^{(e)} \leq E_j \leq r_k^{(e)} \mid B_j = b, i(j) = i_k, \psens, \vec{\theta}) \\
&= p_\text{sens}^{t_+} (1 - p_\text{sens})^{f_-} \left( S_{\vec{\theta}}(l_k^{(e)} - b + 1) - S_{\vec{\theta}}(r_k^{(e)} - b + 2) \right)
\label{imperf-test:eq:ll-ei-lt-ri}
\end{align}

Second, the case when $E_j > r_k^{(e)}$.
In this case, the test at $r_k^{(e)}$ is a false negative, occurring with probability $(1 - p_\text{sens})$.
To avoid having to consider tests after $r_k^{(e)}$, which could greatly complicate the likelihood, I model this case as the episode being right censored at $r_k^{(e)}$.
Taking the same approach as before:
\begin{align}
&\prob(O'(j) = \vec{\nu}_k', E_j > r_k^{(e)} \mid B_j = b, i(j) = i_k, \psens, \vec{\theta}) \\
&= \prob(O'(j) = \vec{\nu}_k' \mid E_j > r_k^{(e)}, B_j = b, i(j) = i_k, \psens, \vec{\theta}) \\
  &\ \ \  \times \prob(E_j > r_k^{(e)} \mid B_j = b, i(j) = i_k, \psens, \vec{\theta}) \\
&= p_\text{sens}^{t_+} (1 - p_\text{sens})^{f_-} (1 - p_\text{sens}) S_{\vec{\theta}}(r_k^{(e)} - b + 2)
\label{imperf-test:eq:ll-ei-gt-ri}
\end{align}

These expressions can now be used to derive $p'_{ik}$.
First, augment the data with $b$, and split into the cases just discussed, omitting the conditioning on $\psens$, $\vec{\theta}$, and $i(j) = i_k$:
\begin{align}
p_{ik}'
=& \prob(O'(j) = \vec{\nu}_k') \\
=& \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \left( \prob(O'(j) = \vec{\nu}_k', E_j \leq r_k^{(e)} \mid B_j = b) + \prob(O'(j) = \vec{\nu}_k, E_j > r_k^{(e)} \mid B_j = b) \right) \prob(B_j = b). \\
\intertext{Now, substitute in \cref{imperf-test:eq:ll-ei-lt-ri,imperf-test:eq:ll-ei-gt-ri} and take out the common factor:}
=\ &  p_\text{sens}^{t_+} (1 - p_\text{sens})^{f_-} \\
 & \times \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \left( S_{\vec{\theta}}(l_k^{(e)} - b + 1) - S_{\vec{\theta}}(r_k^{(e)} - b + 2) + (1 - p_\text{sens}) S_{\vec{\theta}}(r_k^{(e)} - b + 2) \right) \\ 
  & \times \prob(B_j = b \mid p_\text{sens}, \vec{\theta}) \\
=\ &  p_\text{sens}^{t_+} (1 - p_\text{sens})^{f_-} \\
  & \times \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \left( S_{\vec{\theta}}(l_k^{(e)} - b + 1) - p_\text{sens} S_{\vec{\theta}}(r_k^{(e)} - b + 2) \right) \\
  & \times \prob(B_j = b \mid p_\text{sens}, \vec{\theta}).
\label{imperf-test:eq:pia-prime}
\end{align}
Note that if $p_\text{sens} = 1$ then $p_{ik}' = p_{ik}$ (see \cref{perf-test:eq:pia}).

If $\psens$ is fixed (\ie has a point prior) and $\prob(B_j = b \mid \psens, \vec{\theta}) \propto 1$ (as assumed in \cref{E-perf-test}) then:
\begin{align}
p_{ik}'
&\propto \sum_{b = l_k^{(b)}}^{r_k^{(b)}} S_{\vec{\theta}}(l_k^{(e)} - b + 1) - p_\text{sens} S_{\vec{\theta}}(r_k^{(e)} - b + 2).
\label{imperf-test:eq:pia-prime-constant}
\end{align}

@@

This false negative would occur at the first test after the infection episode begins, on day $b_j + \tau_{\sched_{i(j)}}(b_j)$ (see \cref{E-perf-test:sec:prob-undetected}).
A false negative occurring requires that the episode has not yet ended but a negative still occurs.
The episode has not yet ended at the time of the test if $e_j = b_j + d_j - 1 \geq b_j + \tau_{\sched_{i(j)}}(b_j)$, that is the duration of the infection $d_j \geq \tau_{\sched_{i(j)}}(b_j) + 1$.
Conditional on the episode having not yet ended, the test result is negative with probability $1 - \psens$.

For there to be no subsequent positive tests, all tests up until day $e_j$ are false negatives.
I assume there is a negligible probability of two false negatives.
Therefore, this can only occur if the episode ends before another test.
Denote the number of days between $b_j$ and the test following the false negative as $\tau^2_{\sched_{i(j)}}(b_j) \stackrel{\text{def}}{=} \tau_{\sched_{i(j)}}(\tau_{\sched_{i(j)}}(b_j) + 1)$.
Then, the episode ends before this test if $d_j \leq \tau^2_{\sched_{i(j)}}(b_j)$.

Therefore, this mechanism causes episode $j$ to be undetected if all the following conditions hold.
\begin{enumerate}
    \item The episode would have been detected considering only the mechanisms in \cref{E-perf-test:sec:prob-undetected}. That is $\min(\sched_{i(j)}) < b_j \leq T_{i(j)}$ and $e_j \geq \tau_{\sched_{i(j)}}(b_j) + b_j$.
    \item The episode ends in the interval $[\tau_{\sched_{i(j)}}(b_j) + b_j, \tau^2_{\sched_{i(j)}}(b_j) + b_j - 1]$.
      Note that the lower bound here is exactly the bound on $e_j$ in the previous condition.
      Equivalently, $\tau_{\sched_{i(j)}}(b_j) + 1 \leq d_j \leq \tau^2_{\sched_{i(j)}}(b_j)$.
    \item A false negative occurs on day $\tau_{\sched_{i(j)}}(b_j) + b_j$. Conditional on the previous condition, this occurs with probability $1 - \psens$.
\end{enumerate}

The probability of this occurring, conditional on $B_j = b$ where $\min(\sched_{i(j)}) < b \leq T_{i(j)}$ is:
\begin{align}
&\prob \left(
    \tau_{\sched_{i(j)}}(b) + 1 \leq D_j \leq \tau^2_{\sched_{i(j)}}(b)
    \mid B_j = b, \vec{\theta} \right) (1 - \psens) \\
&= \left( S_{\vec{\theta}}(\tau_{\sched_{i(j)}}(b) + 1) - S_{\vec{\theta}}(\tau^2_{\sched_{i(j)}}(b) + 1) \right) (1 - \psens).
\end{align}
Summing over $b$, in the same way as \cref{perf-test:eq:piu}, gives:
\begin{align}
\zeta = (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(\sched_{i(j)}) + 1}^{T_{i(j)}} \left( S_{\vec{\theta}}(\tau_{\sched_{i(j)}}(b) + 1) - S_{\vec{\theta}}(\tau^2_{\sched_{i(j)}}(b) + 1) \right).
\label{imperf-test:eq:zeta}
\end{align}

$p_{iu}'$ is the probability of episode $i$ being undetected, considering both the mechanisms from \cref{E-perf-test} and the new mechanism here.
These mechanisms are mutually exclusive.
Hence, $p_{iu}'$ is the sum of these, $p_{iu}' = p_{iu} + \zeta$.
As in \cref{E-perf-test:sec:prob-undetected}, $1 - p_{iu}'$ is the required quantity.
\begin{align}
1 - p_{iu}'
=& 1 - p_{iu} - \zeta \\
=& \frac{1}{T} \sum_{b = \min \sched_{i} + 1}^{T_{i(j)}} S_{\vec{\theta}}(\tau_{\sched_{i}}(b) + 1) \\
&- (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(\sched_{i(j)}) + 1}^{T_{i(j)}} \left( S_{\vec{\theta}}(\tau_{\sched_{i(j)}}(b) + 1) - S_{\vec{\theta}}(\tau^2_{\sched_{i(j)}}(b) + 1) \right) \\
=& \frac{1}{T} \sum_{b=\min(\sched_{i(j)}) + 1}^{T_{i(j)}} \left( p_\text{sens} S_{\vec{\theta}}(\tau_{\sched_{i(j)}}(b) + 1) + (1 - p_\text{sens}) S_{\vec{\theta}}(\tau^2_{\sched_{i(j)}}(b) + 1)\right).
\label{imperf-test:eq:pit-prime}
\end{align}
